A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-02-25 analysis/paper/paper.Rmd
edit ERGMs and background based on BM's comments

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

<!--limitations of ERGMs-->
However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed “doubly intractable” since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations (details in Supplementary Online Materials) to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on average, 80% of the ties when a third of data is missing.

@@

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as network strategies for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvement in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000], while corporate-based societies stress communal ritual elements, shared power, and wealth differentiation, if any, would be associated with corporate groups, such as age groups [@Siegel1999]. Using the theoretical framework of corporate-network strategies, or horizontal hierarchy, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation ) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete age and sex data (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology.

Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

>>>>

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs are a promising approach for exploration and hypothesis testing of social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

<!--limitations of ERGMs-->
However, those studies also illustrate some limitations of ERGMs, such as sensitivity to missing data and limited ability to incorporate and represent uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the results due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks  [@Caimo2017bayesian]. This is also termed “doubly intractable” since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is to implement ERGMs in a Bayesian framework.

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations (details in Supplementary Online Materials) to better understand dependencies of network variables and overcome computational issues of ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions derived from the context of our data. An important advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information to estimate the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows robust and interpretable uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can accurately predict, on average, 80% of the ties when a third of data is missing.

@@

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of a colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Increased social inequality associated with the use of foreign prestige goods is recognized as a ‘network’ strategy for gaining power [@Blanton1996; @Feinman2000]. Network-based societies are identified in some places in Asia with involvement in long-distance trade [@Carter2015; @Ueda2016; @Liu2006]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!--theoretical framework and research questions-->
The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000]. These may be contrasted with corporate-based societies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age groups, rather than individuals [@Siegel1999]. Using the theoretical framework of corporate-network strategies, we test a hypothesis that changes from a corporate mode (a structure showing more subgroups with less wealth differentiation) to a network mode (a centralized structure with more wealth differentiation) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? and if so, (2) what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent excavation squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). The preservation of human remains at Kiwulan was generally poor, so we have incomplete age and sex data (details in Supplementary Online Materials) and lack health status data. Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on a comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing behaviors before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology.

Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts throughout Taiwan based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that Indigenous people possessing more imported goods were recognized as having higher status in their community [@LiandWu2006].
